Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-28 text/manuscript-full-KS.qmd
update abstract

*The Structure of College Choice* [@RN4982] created "Geomarkets" and the "Market Segment Model" (MSM). Geomarkets carve states and metropolitan areas into smaller geographic units to define local recruiting markets. MSM predicts how student demand for a college varies by Geomarket, based on household socioeconomic characteristics. Geomarkets became an input for two College Board recruiting products. First, the Enrollment Planning Service informs which Geomarkets and high schools a college recruits from. Second, the Student Search Service sells prospective student contact information -- or "student lists" -- which colleges filter by Geomarket to determine which prospect profiles to purchase. Drawing from scholarship on quantification, particularly discussions of correlation and homophily, we conceptualize how recruiting products incorporate Geomarkets. We address two research questions: What is the socioeconomic and racial variation between Geomarkets? How does the socioeconomic and racial composition of included versus excluded prospects vary when student lists filter on Geomarkets? We answer RQ1 by analyzing Census data from 1980, 2000, and 2020. RQ2 analyzes data on student lists purchased by public universities, collected through public records requests. Using a quantitative case study design with metropolitan areas as cases, descriptive statistics and interactive maps suggest Geomarkets mirror deeply-rooted patterns of racial and class segregation. Moreover, student lists that exclude low-income Geomarkets systematically overlook non-White students, first-generation students, and especially first-generation non-White students.

>>>>

The Structure of College Choice (1983) created "Geomarkets" and the "Market Segment Model." Geomarkets carve metropolitan areas into smaller geographies that define local recruiting markets. The Market Segment Model predicts demand for a college in a Geomarket as a function of socioeconomic characteristics. Geomarkets became an input for College Board recruiting products, including their "student list" product, which provides colleges with the contact information of prospective students who meet desired search filter criteria. We draw from scholarship on quantification and address two questions: What is the socioeconomic and racial variation between Geomarkets? How does the socioeconomic and racial composition of included versus excluded prospects vary if student list purchases filter on Geomarkets? We answer RQ1 using Census data and RQ2 using data from actual student list purchases. Geomarket borders are highly correlated with race and class. Student list purchases that exclude low-income Geomarkets systematically overlook high-scoring non-White and first-generation students.

<!-- *The Structure of College Choice* [@RN4982] created "Geomarkets" and the "Market Segment Model" (MSM). Geomarkets carve states and metropolitan areas into smaller geographic units to define local recruiting markets. MSM predicts how student demand for a college varies by Geomarket, based on household socioeconomic characteristics. Geomarkets became an input for two College Board recruiting products. First, the Enrollment Planning Service informs which Geomarkets and high schools a college recruits from. Second, the Student Search Service sells prospective student contact information -- or "student lists" -- which colleges filter by Geomarket to determine which prospect profiles to purchase. Drawing from scholarship on quantification, particularly discussions of correlation and homophily, we conceptualize how recruiting products incorporate Geomarkets. We address two research questions: What is the socioeconomic and racial variation between Geomarkets? How does the socioeconomic and racial composition of included versus excluded prospects vary when student lists filter on Geomarkets? We answer RQ1 by analyzing Census data from 1980, 2000, and 2020. RQ2 analyzes data on student lists purchased by public universities, collected through public records requests. Using a quantitative case study design with metropolitan areas as cases, descriptive statistics and interactive maps suggest Geomarkets mirror deeply-rooted patterns of racial and class segregation. Moreover, student lists that exclude low-income Geomarkets systematically overlook non-White students, first-generation students, and especially first-generation non-White students. -->
